\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{array}
\usepackage[margin=1in]{geometry}

\title{\textbf{GreenPyBench: A Benchmark for Evaluating Automated Python Code Efficiency Optimization}}

\author{
  Miguel Angel Huamani,\quad Aanya Singh Dhankhar,\quad Haoming Qin,\quad Santiago Martinez \\
  University of Illinois Urbana-Champaign \\
  CS 498 AI Agents in the Wild --- Group S11
}

\date{May 5, 2026}

\begin{document}
\maketitle

\begin{abstract}
We introduce \textbf{GreenPyBench}, a reproducible benchmark for evaluating AI agents that optimize Python programs for computational efficiency. Existing code benchmarks emphasize functional correctness but neglect runtime and memory efficiency. GreenPyBench comprises 10 procedural Python tasks spanning common inefficiency patterns: quadratic duplicate detection, linear search, loop aggregation, top-$k$ selection, nested-loop join, string concatenation, generator pipelines, memoization, data-structure choice, and multi-pass fusion. Each task includes a deterministic input generator, a slow reference implementation, an optimized reference implementation, and a PyTest suite. Agents are scored on runtime speedup, peak RAM reduction, and a composite Valid Improvement Score (VIS) that penalizes correctness regressions and code-quality degradation. Any system producing Python source code can be evaluated without modification.
\end{abstract}

%% ============================================================
\section{Introduction}
%% ============================================================

Software energy consumption at the application layer is a largely overlooked source of computational waste~\cite{strubell2019energy,schwartz2020green}. Inefficient Python scripts---featuring $O(n^2)$ loops, non-vectorized operations, redundant copies, and missed memoization---inflate runtime and peak memory in production pipelines, directly increasing energy use. Manual optimization requires profiling expertise and algorithmic reasoning that most developers lack, and existing linters rarely propose complexity-reducing changes.

Existing benchmarks (HumanEval~\cite{chen2021evaluating}, MBPP~\cite{austin2021program}, SWE-Bench~\cite{jimenez2024swebench}) measure correctness, not efficiency. PIE~\cite{shypula2023pie} targets contest-style problems and omits memory metrics and quality penalties. No benchmark provides (a) real-world procedural inefficiency patterns, (b) deterministic profiling harnesses, and (c) joint correctness-and-efficiency scoring with explicit regression penalties.

\textbf{GreenPyBench} fills this gap: it is (1) \emph{reproducible} --- deterministic seeds, fixed package versions, open harness; (2) \emph{correctness-gated} --- efficiency credit withheld for failing rewrites; (3) \emph{agent-agnostic} --- any Python-producing system can be evaluated; and (4) \emph{interpretable} --- per-task and per-family breakdowns reveal which patterns are hardest for agents.

%% ============================================================
\section{Related Work}
%% ============================================================

HumanEval~\cite{chen2021evaluating} and MBPP~\cite{austin2021program} measure function synthesis from docstrings with no efficiency metrics. SWE-Bench~\cite{jimenez2024swebench} targets repository-level bug fixing, also without profiling. PIE~\cite{shypula2023pie} pioneers efficiency as a primary metric but uses competitive-programming problems, excludes memory, and provides no quality penalties. InterCode~\cite{yang2023intercode} evaluates agents in interactive environments but focuses on correctness. GreenPyBench is the first benchmark to target the procedural inefficiency patterns common in scientific Python with a closed-loop profiling harness designed for iterative agents. It operationalizes the Green AI agenda~\cite{schwartz2020green,strubell2019energy} at the application-code layer.

%% ============================================================
\section{Benchmark Design}
%% ============================================================

\subsection{Task Domain and Scope}
GreenPyBench targets 10--50 line procedural Python programs with tractable single bottlenecks. In scope: loop-level optimizations, data-structure substitutions, $O(n^2)\!\to\!O(n)$ reductions, vectorization, and memory streaming. Out of scope: multi-file refactoring, I/O-bound tasks, GPU acceleration, and compiler-level transforms. The harness is purpose-built and does not follow an existing framework.

\subsection{Tasks}
Table~\ref{tab:tasks} summarizes all 10 tasks. Three representative tasks are described below.

\begin{table}[h]
\centering
\caption{GreenPyBench task summary. Difficulty: Easy (E), Medium (M), Hard (H).}
\label{tab:tasks}
\small
\begin{tabular}{clllc}
\toprule
\textbf{\#} & \textbf{Task} & \textbf{Inefficiency} & \textbf{Optimization} & \textbf{Diff.} \\
\midrule
1  & Duplicate Detection   & $O(n^2)$ nested loop       & Hash-based lookup      & M \\
2  & Linear Search         & Full-list scan             & Set membership/bisect  & E \\
3  & Loop Aggregation      & Python loop over array     & NumPy vectorization    & E \\
4  & Top-K Selection       & Full sort for partial res. & Heap-based selection   & M \\
5  & Nested-Loop Join      & $O(n \cdot m)$ cross-join  & Hash join              & H \\
6  & String Concatenation  & Repeated \texttt{+} in loop & \texttt{str.join}     & E \\
7  & Generator Pipeline    & Intermediate list alloc.   & Generator chaining     & M \\
8  & Memoization           & Redundant recursive calls  & LRU cache / DP         & M \\
9  & Data Structure Choice & List where set fits        & Set/dict substitution  & M \\
10 & Multi-Pass Fusion     & Three separate traversals  & Single-pass fusion     & H \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Duplicate Detection (Medium).} Input: $n \in \{10^4,10^5,10^6\}$ integers. Slow: $O(n^2)$ nested comparison. Goal: single-pass frequency table yielding $O(n)$. Output normalized by sorting to ensure test independence.

\paragraph{Nested-Loop Join (Hard).} Input: two tables of $n\!=\!m\!=\!2{,}000$ dicts with a shared key. Slow: $O(n \cdot m)$ double loop. Goal: hash-index one table for $O(n+m)$ join. Output sorted by key before comparison.

\paragraph{Multi-Pass Fusion (Hard).} Input: $n \in \{10^5,10^6\}$ floats. Slow: three separate passes computing sum, mean, std with intermediate allocations. Goal: single-pass running-sum-of-squares fusion. Output matched to six decimal places.

\subsection{Design Principles and Difficulty}
Tasks span five pattern families: algorithmic reduction (1,4,5), data-structure substitution (2,9), vectorization (3), memory streaming (7,10), and caching (8). All slow implementations are idiomatic Python a competent developer might write, not adversarially obfuscated code. Correctness tests normalize outputs before comparison. Table~\ref{tab:difficulty} shows the difficulty distribution; harder tasks offer larger optimization headroom, validating the labels.

\begin{table}[h]
\centering
\caption{Difficulty distribution. Avg.\ speedup is reference optimized vs.\ slow.}
\label{tab:difficulty}
\small
\begin{tabular}{lccl}
\toprule
\textbf{Level} & \textbf{Count} & \textbf{Avg.\ Speedup} & \textbf{Tasks} \\
\midrule
Easy   & 3 & 4.1$\times$  & 2, 3, 6 \\
Medium & 5 & 11.3$\times$ & 1, 4, 7, 8, 9 \\
Hard   & 2 & 38.7$\times$ & 5, 10 \\
\bottomrule
\end{tabular}
\end{table}

%% ============================================================
\section{Evaluation Protocol}
%% ============================================================

\paragraph{Metrics.} Per task: (1) \emph{Correctness pass rate} over PyTest suite; (2) \emph{Runtime speedup} (\%) --- median of 7 measured runs (after 2 warm-ups) in an isolated subprocess; (3) \emph{Peak RAM reduction} (\%) --- median \texttt{tracemalloc} snapshot.

\paragraph{Scoring.} Correctness is a hard gate. The per-task Valid Improvement Score (VIS) is:
\begin{equation}
\text{VIS} = \max\!\left(0,\; 0.7\times\text{runtime\_gain} + 0.3\times\text{memory\_gain} - \text{quality\_penalty}\right)
\end{equation}
Gains are clipped at zero. $\text{quality\_penalty}=0.25$ if cyclomatic complexity increases $>$20\% or lines of code increase $>$30\% without achieving $\geq$10\% speedup or $\geq$5\% RAM reduction (measured via \texttt{radon}).

\paragraph{Aggregation and Reporting.} Report mean correctness, mean runtime speedup, mean RAM reduction, mean VIS, and regression rate---broken down by difficulty tier and pattern family. LLM-based methods should be run at least three times to estimate variance. Submissions must specify model name, temperature, iteration limit, and Python environment.

%% ============================================================
\section{Benchmark Validation}
%% ============================================================

\paragraph{Harness reliability.} Running each slow implementation 20 times yielded CV $<5\%$ for runtime and CV $<3\%$ for RAM across all tasks, confirming that 7-run median aggregation produces stable estimates.

\paragraph{Baseline results.} Table~\ref{tab:baseline} reports four conditions. The static rule baseline applies five deterministic rewrites (\texttt{join}, \texttt{@lru\_cache}, NumPy, \texttt{set()}, \texttt{heapq}) and returns the original when no rule fires. It achieves non-trivial speedups on exactly three tasks --- String Concatenation (98.8\%, VIS 69.4), Memoization (100.0\%, VIS 70.0), Loop Aggregation (75.4\%, VIS 52.8) --- and 0\% on the remaining seven, confirming that the benchmark requires genuine reasoning beyond rule matching. Full LLM-based results are pending API completion and marked \textbf{[TBD]}.

\begin{table}[h]
\centering
\caption{Baseline results on GreenPyBench (10 tasks). RT = runtime speedup, RAM = peak RAM reduction, both means over passing tasks. \textbf{[TBD]} = pending evaluation.}
\label{tab:baseline}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{System} & \textbf{Correct (\%)} & \textbf{RT Speedup (\%)} & \textbf{RAM Red.\ (\%)} & \textbf{VIS} \\
\midrule
Original slow code        & 100 & 0.0  & 0.0  & 0.00 \\
Static rule baseline      & 100 & 18.4 & 6.1  & 0.14 \\
Single-shot LLM           & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
One-pass profile-guided   & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} & \textbf{[TBD]} \\
Reference optimized impl. & 100 & 82.1 & 41.6 & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Human baseline.}
% TODO: Have 2-3 participants manually optimize Tasks 1, 5, 8 without tools.
% Measure with harness.py (7-run median). Fill in table below and uncomment.
%
% \begin{table}[h]
% \centering
% \caption{Human baseline (n=3 participants) on three representative tasks.}
% \small
% \begin{tabular}{lccc}
% \toprule
% Task & Speedup (\%) & Correct (\%) & Time (min) \\
% \midrule
% Duplicate Detection & [TBD] & [TBD] & [TBD] \\
% Nested-Loop Join    & [TBD] & [TBD] & [TBD] \\
% Memoization         & [TBD] & [TBD] & [TBD] \\
% \bottomrule
% \end{tabular}
% \end{table}
\textbf{[TBD]} Human baseline pending. Three participants will manually optimize Tasks 1, 5, and 8 without tool assistance; speedups will be measured using the standard harness and reported in the final submission.

%% ============================================================
\section{Conclusion}
%% ============================================================

GreenPyBench is the first reproducible benchmark for AI agents on Python code efficiency, with joint correctness-and-efficiency metrics, regression penalties, and a profiling harness for iterative systems. Static rule baselines confirm the benchmark requires genuine algorithmic reasoning. Full results will be reported upon completion of LLM evaluation runs. The harness, tasks, seeds, and baselines will be released to support reproducible Green AI research. Future extensions include 20 tasks, energy measurement via RAPL counters, and evaluation on real-world scientific codebases.

\paragraph{Changelog.} Final version adds confirmed per-task static rule VIS results, refines \texttt{tracemalloc} harness details, and expands task design principles. Human baseline and LLM rows in Table~3 pending completion.

%% ============================================================
\begin{thebibliography}{10}
\bibitem{austin2021program} J. Austin et al. Program synthesis with large language models. \emph{arXiv:2108.07732}, 2021.
\bibitem{chen2021evaluating} M. Chen et al. Evaluating large language models trained on code. \emph{arXiv:2107.03374}, 2021.
\bibitem{jimenez2024swebench} C.E. Jimenez et al. SWE-bench: Can language models resolve real-world GitHub issues? \emph{arXiv:2310.06770}, 2024.
\bibitem{madaan2023chatgpt} A. Madaan et al. Is your code generated by ChatGPT really correct? \emph{arXiv:2305.01210}, 2023.
\bibitem{schwartz2020green} R. Schwartz et al. Green AI. \emph{Commun.\ ACM}, 63(12):54--63, 2020.
\bibitem{shypula2023pie} A. Shypula et al. Learning performance-improving code edits. \emph{arXiv:2302.07867}, 2023.
\bibitem{strubell2019energy} E. Strubell et al. Energy and policy considerations for deep learning in NLP. \emph{arXiv:1906.02629}, 2019.
\bibitem{xia2023evoeval} C.S. Xia et al. EvoEval: Evolving coding benchmarks via LLMs. \emph{arXiv:2403.19114}, 2023.
\bibitem{yang2023intercode} J. Yang et al. InterCode: Standardizing and benchmarking interactive coding. \emph{arXiv:2306.14898}, 2023.
\end{thebibliography}

\end{document}